### 2020-08-17 response/response-demography-3.tex

\begin{document}

\begin{center} \textbf{\large Birth Spacing in the Presence of Son
Preference and 
Sex-Selective Abortions: India's Experience over Four Decades}
\end{center}

\begin{center} Response to Editor and Reviewer Comments \end{center}

\noindent I have attached the revised version of my paper,
``Birth Spacing in the Presence of Son Preference and Sex-Selective
Abortions: India's Experience over Four Decades.''
Thank you for the comments and suggestions. 
Below I outline my response to each issue raised.

As part of rewriting the paper in response to the comments, I have changed several 
additional components to ensure that it is easier to follow and understand.
The following are the more substantial changes not covered below in my responses:
\begin{enumerate}
    \item I substantially rewrote the Abstract to directly discuss the three main 
    questions I address in the paper
	\item ``Estimation Strategy'' section changes:
	\begin{enumerate}
		\item Clarify proportionality assumption
		\item Better explain the set-up of the model and transformation of data needed
		\item Remove discussion of explanatory variables, so they are
		only covered once (in the ``Data'' section)
		\item Better explain how I calculate measures used for the spacing
		analysis and all explanations gathered in this section
		\item Remove the formula for percent boys since it added little. 
		I have added a more formal version of the model to the online Appendix in the
		``Empirical Model Details'' section.
	\end{enumerate}
	\item ``Data'' section changes:
	\begin{enumerate}
		\item I now lead with a discussion of which birth spells I analyze right after 
		introducing data
		\item Move discussion of restriction to only Hindus to be the first since that is 
		the most important. 
		See also my response to Reviewer 2 below.
	\end{enumerate}
	\item Consistent presentation of results:
	\begin{enumerate}
		\item Update table notes to describe calculation methods more clearly
	\end{enumerate}
\end{enumerate}

Finally, I have decided to amend the paper title slightly to convey the main questions 
addressed.
The new title is ``Birth Spacing and Fertility in the Presence of Son Preference and 
Sex-Selective Abortions: India's Experience over Four Decades.'' 
If this becomes too long, removing ``Son Preference'' would be a possibility.
Please let me know if you have any strong opinions on this.

@@

\begin{enumerate}

\item We note that you kept to the table/figure limit; however, some of
the figures are several pages long and relatively small. These will have
to be evaluated at the copy edit stage as there is no obvious quick fix
- the figures reflect the analyses carried out (and some additional ones
are in an online appendix). Please make sure to submit as
high-resolution image as possible in the format required by the journal.
We are somewhat concerned about figure/table size within the print form
of the journal.

[Response:]

After trying a substantial number of different approaches, I believe the best way to solve 
this problem is a complete rewrite of the birth spacing analysis section.
Therefore, I now present three percentile birth intervals---25th, median, and 75th---together
with the sex ratio and the parity progression probability.
Furthermore, to keep within the figure/table limit and not have multiple-page figures,
I moved the results for urban women without education and rural women with twelve or more 
years of education to the online Appendix.

The main advantage of this new approach---besides making the figures less challenging to 
read---is that it more clearly ties the section directly to the main question I am asking.
Furthermore, the new set-up makes it easier to understand similarities and differences 
across groups and compare the results to the prior literature.
It also ties the birth spacing discussion directly to the fertility section that follows.

The rewritten section has three parts.
I now first present the broad trends for parity progression and sex ratio to set the
stage for the analysis.

Next, I discuss how birth spacing has changed in situations where there is the least use
of sex selection (less educated women no matter the sex composition and among 
better-educated women with one or more sons).
To this part, I added a discussion of how my results compare with the prior literature 
\citep{Rutstein2011,Casterline2016}.
To tie the results to the labor force participation, I also show how average spacing
has increased the most for the women least likely to work.
Finally, I show how the longest spacing has changed with declining fertility, and discuss 
why the behavior of the best-educated women appear to be different from other groups in 
India and elsewhere.

The last part examines how birth spacing changed with the use of sex selection.
To illustrate, I use the third spell for the best-educated women, especially the very 
rapid increase right after sex selection became available.
I then show where the reversal of the traditional spacing pattern occurred.
As part of my response to reviewer 2's question about the period grouping, I include a 
discussion of whether the predicted decline in the use of sex selection came about.
I end with a summary of results to tie the section to the fertility section.

Finally, I can produce figures in whatever format is best for printing.
I can also easily change things like line width and labeling if that helps with the
printing process.

@@

\begin{enumerate}

\item The overall topic and research question is not clear in the
introduction. The main topic relates to changes in birth spacing over
time, which are influenced by sex selection/sex composition and maternal
education, and how this affects fertility and infant mortality. This is
a long list of interesting questions and analyses but this is confusing
for the reader. It would be useful to have a succinct statement about
what the study is about (what are key independent variables/concepts and
key outcomes) and place that in the introduction and before the data and
analyses section. This way, the reader can understand why the analyses
will proceed in a particular order and what is being tested. In
addition, the titles of the figures and tables should be reviewed to see
if they could better reflect the analyses undertaken. Overall, the
reader needs to be reminded at the beginning and in each section what
you aim to do.

[Response:]

I have rewritten the Introduction to emphasize the main question I address (``how birth 
spacing responded to these changes, especially the spread of sex selection'').
I follow this with the two questions addressing the effects of changes in 
birth spacing (``to what extent do changes in birth spacing bias our standard fertility 
estimates?'' and `` how do the changes in birth spacing and the increased use of abortions 
impact children's chances of survival?'').

I next outline how the study progresses, focusing on how I examine each of the three 
questions, including the key outcomes and key explanatory variables. 
To prevent exceeding the word count, I have edited the entire Introduction.

The beginning of each section now highlights what I question I am addressing.

Furthermore, I have tried to removed repetitions throughout the paper. 
One example is the discussion of tempo effects and mortality, where the motivation is now 
only in the Introduction rather than being spread over multiple sections.

As described above, I have completely redone the discussion of how spacing changed.
I hope this will make it easier for the reader to see the answers to the paper's central 
question.

Finally, I have edited the figure and table titles to reflect the analyses better.

@@

I now define ``sex selection'' as ``the selective abortion of female fetuses 
based on prenatal sex determination'' in the second paragraph of the paper.

\item Please provide a motivation for using maternal education as a key
stratification measure and justify education cutoffs (which have meaning
in India).

[Response:]

To address this, and the questions about theories of female labor supply below, I have
combined and reorganized the background and conceptual framework sections.
I begin this combined section with a discussion of the three main reasons for using 
maternal education as a key stratification measure.
First, higher female education is associated with lower fertility, and, thereby, the 
potential increased use of sex selection. 
Second, female labor force participation in India, and other developing countries, 
tend first to decrease and then increase with education. 
Finally, child morbidity and mortality decrease the better educated the mother.

After discussing why female education is a crucial variable, I explain the choice of 
cut-offs, which I based on the prior literature on fertility and son preference in India 
and the NFHS reports.
The exception to the cut-offs in the NFHS reports is that I combine
the <5 years and 5--7 years of schooling completed and 
the 8--9 and 10--11 years to reach sufficient sample sizes.

This grouping also allows me to capture the U-shape in female labor force participation,
which I refer to in my discussion of the changes in birth spacing when sex selection
is less used.
In principle, I could split the upper level of education into having graduated secondary 
and having graduated with a university degree since women with a completed university 
degree have a higher labor force participation. 
The issue is that the sample of university graduates is too small to allow for estimations 
for the first three rounds of the surveys. 
Furthermore, with four groups, there is already a substantial number of results to present.

@@

\item The word ``spell'' is introduced with no definition. Is spell the
same as birth interval? Is it measured in months? This information would
be useful for the reader.

[Response:]

I now define spell as the unit of analysis---the period from one parity birth to 
the following birth or censoring---when I first use the word in the ``Estimation 
Strategy'' section.
Furthermore, I clarify that for estimation purposes, spells begin nine months 
after the previous birth since this is the earliest we should expect to observe 
a new birth.
% This duration is equivalent to the interpregnancy interval, which is often used in 
% the literature on spacing and child mortality.

To make comparisons with both the NFHS reports and the prior literature on
birth spacing easier and avoid confusion, I have rewritten the discussions of 
spacing and redone the graphs and table, so all results now reflect birth 
intervals in months.
I still retain the use of ``spell'' to indicate a generic period from one parity 
birth to another---i.e.,the second spell is from first birth to the second birth or
censoring---in the graphs/tables.

In the Data section, I have also changed the discussion of the imposed censoring 
clearer by specifying the cut-offs for both the spell duration and the birth interval.
In the ``Estimation Strategy'' section, I explain that I add nine months
to the estimated spell length to get to birth intervals.
In the ``Mortality and the Changing Birth Spacing'' section, I retained the
definition of the periods but updated the numbers and the discussion of
results. 
Hence, they now reflect birth intervals rather than pregnancy intervals.

@@

\item Please include a brief discussion or footnote about the quality of
birth date data in the DHS/NSFH and how this could bias your results or
not.

[Response:]

I have expanded the discussion of the quality of the birth history data in the ``Data'' 
section and its importance. 
As part of this expanded discussion, I included more information about systematic recall error, why it is likely the main concern here, the effect on the estimates, and how I address this problem.
I have also included a reference to \citet{Schoumaker2014} that finds that the first
rounds of the NFHS are of ``moderate quality,'' although that is without the correction
for recall error that I use.

@@

\item Perhaps the conclusion could start with a brief overview of your
topic and research question again to remind the reader of the overall
motivation for the study.

[Response:]

I have added a brief synopsis of the changes in India in the beginning of
the Conclusion and recap the three questions after that.
I have also edited the Conclusion to tie better with the questions.

@@

Yes, I already had it listed as World Bank Economic Review 2015.
Was there maybe another paper that I should cite?

\item Please have the paper carefully reviewed/edited, as there are
numerous grammatical and syntax mistakes.

[Response:]

As part of the rewrite, I have gone through the paper multiple times to try to catch
any remaining mistakes.
I have also had an copy editor go through the paper.

@@

\section*{Reviewer 1}

I already liked the previous version of this paper. I think the author
has done a good job incorporating most of my comments and those of the
other reviewers. Even though I would sometimes have made different
choices (for example, including fewer figures), the choices the author
has made are well understandable and appropriately motivated. I think
the paper is a useful contribution to Demography and my advice is to
accept it.

[Response:]

I have tried to limit the number of figures as detailed above in my response to
the Editors.

@@

\newpage

\section*{Reviewer 2}

% [With respect to the reviewer’s comments, please address comments #1, #2, and #4.]

@@

\item There is actually no data on sex selection and in a sense the data
is backed out using evidence on birth spacing, institutional information
on the availability of technology that enables sex selection (ultra
sound techniques) and government regulations. To do this the author
subdivides the time into four segments: 1972--1984, 1985--1994, 1995--2004
and 2005--2016. I understand the logic of the first two breaks 1984 and
1994, but what is the logic of the third break in 2004? This needs to be
made clearer.

[Response:]

The 2004 break is an attempt to understand whether we are observing the beginning of
a reversal in son preference and the use of sex selection as hypothesized by part of
the prior literature.
In the ``Data'' section, I have split the paragraph on the periods into two and expanded 
the discussion of the motivation for the breaks, including adding two additional
references arguing that we might see the reversal.
Clearly, the 2004/05 break is somewhat arbitrary, but does have the advantage of making 
the periods close to the same length (there are relatively few birth intervals that begin 
in 2015 and 2016).

I now also discuss the changes between the last two papers in more detail in the 
subsection on birth spacing with sex selection and in the Conclusion.

@@

\item While I like results being presented in figures, what I don’t like
is that no standard errors or confidence intervals are presented.
Therefore, it is difficult to figure out whether the differences between
categories (composition of existing children) are statistically
significant or not.

[Response:]

The main reason for not showing confidence intervals is that the figures would become
close to impossible to read and previous reviewers felt that the tables were too dense
to understand easily, a point I agree with.
The original tables do have standard errors and test of whether differences are 
statistically significant and these are available in the online Appendix.
I have made it clearer in the text that the underlying values for the graphs and the
bootstrapped standard errors are available in the Appendix.

@@

The discussion of labor supply was disjointed and unclear because I had split the 
background and conceptual framework sections, making it difficult to see the connection.
As explained above, I have combined and reorganized those two sections into one and
removed the Goldin theory discussion since it is not directly relevant to my main
argument of why labor supply is relevant.

\item Overall, I felt that the paper is still quite densely written and
it is not easy to follow what is going on. Considerable time needs to be
spent re-writing the paper and making it easier to read.

[Response:]

As I have outline above, I have reorganize large parts of the paper and significantly
rewritten the whole paper to make it easier to follow.

### 2020-08-19 paper/sex_selection_spacing-ver4.tex
Edit based on proof reader comments

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

% Demography abstract
\noindent 
In India, the most substantial lengthening in Hindu women's birth intervals over the last 
four decades occurred for the women most likely to use sex-selective abortions. 
For well-educated women with two girls, the median birth interval to the third child 
increased by nearly 15 months---and the 75th percentile interval by 21 months---as the 
sex ratio approached 80\% boys. 
Because of the rapid lengthening in spacing, standard fertility rates significantly 
overestimated how fast cohort fertility fell. 
Despite convergence, cohort fertility is still 10 to 20\% higher than the 
fertility rate, and above replacement level for all but the best-educated urban women. 
With the lengthening from sex selection, some women show a reversal of the traditional 
spacing pattern; instead of the shortest birth interval in the absence of sons, they now 
have the longest. 
Finally, infant mortality has declined over time for all groups, but the decline was the 
fastest for the less educated. 
Short birth spacing is still associated with higher mortality, although considerably less 
so for the best-educated women. 
There is no evidence that the increasing use of sex selection is associated with higher 
infant mortality. 
Neither does it appear that the use of sex selection is declining.

@@

\noindent Keywords: India, prenatal sex determination, censoring, competing risk, 
non-proportional hazard
\end{abstract}

@@

% Background
India has experienced many positive changes over the past four decades:
the economy has grown substantially,
educational attainment has increased for males and females,
and the total fertility rate has fallen to 2.2 children
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

@@

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, prior research has failed to appreciate that each abortion increases the
interval between births by 6 to 12 months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
although we do not know by how much. 
Second, greater educational attainment by women, higher household income, and the low and 
declining female labor force participation all likely influence birth spacing. 
Thus, the combined changes in birth spacing may outpace what we have observed in 
other countries.

@@

First, did changes in birth spacing bias the fertility estimates for India? 
With longer spacing, mothers will be older at each parity, and this tempo effect makes the 
total fertility rate a downward biased estimate of cohort fertility
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}. 
Hence, if birth intervals increased substantially, then India's fertility may be higher 
than generally accepted.

@@

% How the study progresses
To investigate how birth spacing has changed, I use a competing risk hazard model with 
two exit states: the birth of a girl or the birth of a boy. 
I apply the model to the birth histories of Hindu women covering 1972 to 2016 using 
data from the four National Family and Health Surveys (NFHS).

@@

First, birth intervals lengthen over the four decades---more so the higher the parity, the 
more educated the woman, and the higher the percentile. 
Women most likely to use sex selection---well-educated women with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some 
women with no sons having \emph{longer} birth intervals than those with sons. 
Those least likely to use sex selection---less-educated women in rural areas---still follow the 
traditional pattern of short spacing when they have girls with only some evidence of sex 
selection. 
The likelihood of a very short birth interval changed little.

Second, the fertility rate substantially overestimated how fast cohort fertility fell in the 
1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still 10 to 
20\% higher than the fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

@@

I divide eduction into four groups: no education, one to seven years, 8 to 11 years, and 
12 and more years.
The latter two correspond to having completed primary and completed secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and 11 and 12 for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and five to seven years of schooling completed and the eight to nine 
and 10 to 11 years of schooling completed.
This grouping allows me to capture the differences across education levels discussed below, 
while also having groups large enough for the empirical method.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
twenty years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90\% for the 1930s cohorts to 
less than 20\% for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than 60\% for the 1990s cohorts, with about half in the 8 to 11 group and the 
other half in the 12 plus group.
In urban areas, the proportions with no education or one to seven years have each declined 
to just below 10\%.
Most of the increase in urban female education came from the 12 plus group, which 
now account for more than half of all urban women for the lastest cohorts.

\begin{figure}[htpb]
\centering
\subfloat[Rural]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
}
\caption{Distribution of education by cohort for women twenty years or older at survey}
\label{fig:education_over_time}
\end{figure}

@@

However, even as the level of female education has increased, the female labor force 
participation in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
India's female labor force participation is now lower than most other countries and does 
not yet show any signs of increasing \citep{Klasen2015,Chatterjee2018}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with the highest percent working for women with 
either no education or twelve or more years and the lowest for women with eight to 
eleven years of education.%
\footnote{
See Appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

The low and declining female labor force participation, especially for younger women, 
suggests that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak for this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to own wages \citep{Bhargava2018}.

@@

The introduction of sex selection allows parents to avoid giving birth to girls but 
increases the expected duration to the next birth.
Theory suggests that sex selection increases with lower desired fertility and with
higher parity for a given desired number of children \citep{Portner2015b}.
Sex selection is more widespread among better-educated than less-educated and among urban 
than rural women, which is consistent with lower desired fertility increasing the
use of sex selection 
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
lowers the relative costs of using sex selection and have long birth intervals.

The combination of rising incomes and continued son preference may lead to even longer 
spacing than expected from the income effect alone.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and the increasing income potential for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.

If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals \citep{Zajonc1975,Zajonc1976,Razin1980}.
However, the evidence on spacing's effect on child quality measures such as IQ 
and education is mixed for developed countries and non-existing for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is health and mortality, where longer spacing does lead to better outcomes, 
although this relationship weakens with maternal education 
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

@@

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the associated change in the composition of better-educated women 
is that this group's behavior would change.
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, they adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation suggests that this process still operates.

@@

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better-educated 
because their household income increases the most---even with declining female labor 
participation---and their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change 
the use of sex selection.

@@

% Requirements for method and the intuition behind each:  
%  - Account for censoring (hazard model)
%  - Work both with and without sex selection (competing risk part)
%  - Allow for changes in use of sex selection within spell (flexible specification of baseline hazard ?)
%  - Capture that the shape of the hazard function differ across groups
%    (non-proportionality); needs a better description, maybe using example
%    of shorter spacing and/or use of sex selection. Differences in parity
%    progression likelihood.

@@

First and foremost, the use of sex selection means that the sex of the next child is 
no longer random and that the spacing to the birth of a boy will differ from the 
spacing to a girl. 
Therefore, I use a competing risk set-up that captures both the non-randomness of 
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

>>>>

\title{Birth Spacing and Fertility in the Presence of Son Preference and Sex-Selective Abortions:
India's Experience Over Four Decades%
\protect\thanks{%
I am grateful to Andrew Foster and Darryl Holman for discussions about the method.
I owe thanks to Shelly Lundberg, Daniel Rees, David Ribar, 
Hendrik Wolff, seminar participants at University of Copenhagen, University of Michigan, 
University of Washington, University of Aarhus, the Fourth 
Annual Conference on Population, Reproductive Health, 
and Economic Development, and the Economic Demography Workshop for helpful 
suggestions and comments.
I would also like to thank Nalina Varanasi for research assistance.
Support for development of the method from the University of Washington Royalty 
Research Fund and the Development Research Group of the World Bank is gratefully 
acknowledged.
The views and findings expressed here are those of the author and
should not be attributed to the World Bank or any of its member countries.
Partial support for this research came from a Eunice Kennedy Shriver National
Institute of Child Health and Human Development research infrastructure grant,
5R24HD042828, to the Center for Studies in Demography and Ecology at the
University of Washington.
}
}

@@

% Demography abstract
\noindent 
In India, the most substantial lengthening in Hindu women's birth intervals over the past 
four decades has occurred for the women most likely to use sex-selective abortions. 
For well-educated women with two girls, the median birth interval before the third child 
has increased by nearly 15 months---and the 75th percentile interval by 21 months---as the 
sex ratio has approached 80\% boys. 
Because of the rapid lengthening in spacing, standard fertility rates have significantly 
overestimated how fast cohort fertility fell. 
Despite convergence, cohort fertility is still 10\%--20\% higher than the 
fertility rate and above replacement level for all but the best-educated urban women. 
With the lengthening from sex selection, some women show a reversal of the traditional 
spacing pattern; instead of the shortest birth interval in the absence of sons, they now 
have the longest. 
Finally, infant mortality has declined over time for all groups, but the decline was 
fastest for the less educated. 
Short birth spacing is still associated with higher mortality, although considerably less 
so for the best-educated women. 
There is no evidence that the increasing use of sex selection is associated with higher 
infant mortality. 
Neither does it appear that the use of sex selection is declining.

@@

\noindent Keywords: India, prenatal sex determination, censoring, competing risk, 
nonproportional hazard
\end{abstract}

@@

% Background
India has experienced many positive changes over the past four decades:
The economy has grown substantially,
educational attainment has increased for males and females,
and the total fertility rate has fallen to 2.2 children
\citep{Bosworth2008,Dharmalingam2014,
International-Institute-for-Population-Sciences-IIPS2017}.

@@

% Questions addressed and why valuable
The main question I address here is how birth spacing responded to these changes, 
especially the spread of sex selection. 
The motivation is twofold.
First, researchers have failed to appreciate that each abortion increases the
interval between births by 6--12 months.%
\footnote{
The increase consists of three parts. 
First, after an abortion, the uterus needs at least two menstrual cycles to recover, 
or the likelihood of spontaneous abortion increases substantially \citep{zhou00b}. 
Second, the waiting time to conception is one to six months. 
Finally, sex determination tests are reliable only from three months of gestation. 
}
Hence, the growing use of sex-selective abortions may substantially increase birth spacing, 
although we do not know to what extent. 
Second, greater educational attainment by women, higher household income, and the low and 
declining female labor force participation all likely influence birth spacing. 
Thus, the combined changes in birth spacing may outpace what we have observed in 
other countries.

@@

First, did changes in birth spacing bias the fertility estimates for India? 
With longer spacing, mothers will be older at each parity, and this tempo effect makes the 
total fertility rate a downward-biased estimate of cohort fertility
\citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}. 
Hence, if birth intervals increased substantially, then India's fertility may be higher 
than generally accepted.

@@

% How the study progresses
To investigate how birth spacing has changed, I use a competing risk hazard model with 
two exit states: The birth of a girl or the birth of a boy. 
I apply the model to the birth histories of Hindu women covering 1972--2016 using 
data from the four National Family and Health Surveys (NFHS).

@@

First, birth intervals lengthened over the four decades, and the lengthening was longer, 
the higher the parity, the more educated the woman, and the higher the percentile. 
Women most likely to use sex selection---well-educated women with no sons---had the most 
substantial lengthening of birth intervals and the most biased sex ratios. 
As a result, we now see cases that reverse the traditional spacing pattern, with some 
women with no sons having \emph{longer} birth intervals than those with sons. 
Those least likely to use sex selection---less-educated women in rural areas---still follow the 
traditional pattern of short spacing when they have girls, with only some evidence of sex 
selection. 
The likelihood of a very short birth interval changed little.

Second, the fertility rate substantially overestimated how fast cohort fertility fell in the 
1990s and early 2000s as spacing began to increase. 
Although the two have lately been converging, the predicted cohort fertility is still 
10\%--20\% higher than the fertility rate. 
Furthermore, predicted cohort fertility is still at or above replacement level for all but 
the best-educated urban women.

@@

I divide education levels into four groups: No education, 1--7 years, 8--11 years, and 
12 and more years.
The latter two correspond to having completed primary and secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and 11 and 12 for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and 5--7 years of schooling completed and the 8--9 
and 10--11 years of schooling completed.
This grouping allows me to capture the differences across education levels discussed below, 
while also having groups large enough for the empirical method.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
20 years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90\% for the 1930s cohorts to 
less than 20\% for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than 60\% for the 1990s cohorts, with about half in the 8--11 group and the 
other half in the 12 plus group.
In urban areas, the proportions with no education or 1--7 years have each declined 
to just below 10\%.
Most of the increase in urban female education came from the 12 plus group, which 
now accounts for more than half of all urban women for the most recent cohorts.

\begin{figure}[htpb]
\centering
\subfloat[Rural]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_rural}
    \end{minipage}
} 
\subfloat[Urban]{
    \begin{minipage}{0.49\textwidth}
        \includegraphics[width=\textwidth]{educ_over_time_urban} 
    \end{minipage}
}
\caption{Distribution of education by cohort for women 20 years or older at survey}
\label{fig:education_over_time}
\end{figure}

@@

However, even as the level of female education has increased, the female labor force 
participation in both urban and rural areas has stagnated or decreased
\citep{Klasen2015,Fletcher2017,Afridi2018,Bhargava2018,Chatterjee2018,Bhargava2019}.
India's female labor force participation is now lower than that of most other countries and does 
not yet show any signs of increasing \citep{Klasen2015,Chatterjee2018}.
In line with previous research, the NFHS data show a U-shaped relationship between 
education and working for married women, with the highest percentage working being women with 
either no education or 12 or more years and the lowest being women with 8--11 years of education.%
\footnote{
See appendix Figures \ref{fig:work_by_survey} through \ref{fig:work_family_by_survey}. 
}

The low and declining female labor force participation, especially for younger women, 
suggests that families face little incentive to space children more closely 
together for economic reasons.
One explanation is that household income has increased so substantially that the income 
effect dominates any substitution effect.
Two findings speak to this effect.
First, although real wages for both men and women have almost doubled between 1987 and 
2011, the mean male wage is still close to 70\% higher than the female wage 
\citep{Klasen2015,Bhargava2018}.
Second, women’s labor supply appears to be more negatively elastic to husbands' 
wages than positively elastic to their own wages \citep{Bhargava2018}.

@@

The introduction of sex selection allows parents to avoid giving birth to girls but 
increases the expected interval to the next birth.
Theory suggests that sex selection increases with lower desired fertility and with
higher parity for a given desired number of children \citep{Portner2015b}.
Sex selection is more widespread among better-educated than less-educated and among urban 
than rural women, which is consistent with lower desired fertility increasing the
use of sex selection 
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
lowers the relative costs of using sex selection and having long birth intervals.

The combination of rising incomes and continued son preference may lead to even longer 
spacing than might be expected from the income effect alone.
As women's education increases, their productivity in the production of offspring human 
capital also increases.
With relatively more boys born because of increased access to sex-selective 
abortions and the increasing income potential for (male) offspring, demand for 
better-educated women can increase, even if they do not participate in the labor market 
\citep{Behrman1999}.

If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals \citep{Zajonc1975,Zajonc1976,Razin1980}.
However, the evidence on spacing's effect on child quality measures such as IQ 
and education is mixed for developed countries and nonexisting for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is health and mortality, where longer spacing does lead to better outcomes, 
although this relationship weakens with maternal education 
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

@@

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the associated change in the composition of better-educated women 
is that this group's behavior would change.
However, ``Sanskritization'' implies that as lower-castes females gain access to 
education and their husbands' income increases, they adopt higher-caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation suggests that this process still operates.

@@

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better educated 
because their household income increases the most---even with declining female labor 
participation---and because of their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change 
the use of sex selection.

@@

% Requirements for method and the intuition behind each:  
%  - Account for censoring (hazard model)
%  - Work both with and without sex selection (competing risk part)
%  - Allow for changes in use of sex selection within spell (flexible specification of baseline hazard ?)
%  - Capture that the shape of the hazard function differ across groups
%    (nonproportionality); needs a better description, maybe using example
%    of shorter spacing and/or use of sex selection. Differences in parity
%    progression likelihood.

@@

First and foremost, the use of sex selection means that the sex of the next child is 
no longer random and that the spacing to the birth of a boy will differ from the 
spacing to a girl. 
Therefore, I use a competing risk setup that captures both the non-randomness of 
the birth outcome and the differential spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there was underreporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, even without sex selection, it is unlikely that characteristics, such as 
the sex composition of previous births, have the same effects throughout the entire 
birth interval. 
The proportional hazard model requires that the hazard for any individual is a 
fixed proportion of the hazard for any other individual. 
Nonconstant effects violate that assumption, and the results from a proportional hazard 
model would, therefore, be biased. 
The proportionality assumption is especially problematic for higher-order birth 
intervals because there are substantial differences across groups in the likelihood 
of progressing to the next birth and how soon couples want their next child if they 
are going to have one \citep{Whitworth2002,Bhalotra2008,Kim2010}.
